---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ipolygrowth

<!-- badges: start -->
<!-- badges: end -->

The goal of ipolygrowth is to calculate bacterial growth curve parameters using fourth degree polynomial functions. Functions are available for a single biological sample or multiple samples.


## Installation

The package can be installed from CRAN with:

```{r, eval=FALSE}
install.packages("ipolygrowth")
```

or the development version from [GitHub](https://github.com/) with:

``` {r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("https://github.com/kivanvan/ipolygrowth", upgrade = F, quiet = T)
```


## Example

The example data comes from the [growthrates](https://CRAN.R-project.org/package=growthrates) package.  
This is a basic example which shows you how to use the single sample function:  

```{r example, message=FALSE}
library(ipolygrowth)
library(dplyr)

# example data comes from the growthrates package (available on CRAN)
data <- growthrates::bactgrowth

# subset data to a single biological sample
df.singlesample <- data %>% dplyr::filter(strain == "D", conc == 0)

# calculate growth curve parameters using ipolygrowth function
out.singlesample <- ipg_singlesample(data = df.singlesample, time.name = "time", y.name = "value")
```

The output is a list, including a table of growth parameter estimates, the polynomial model, a table of beta coefficients, and a table of fitted values. Growth parameters include peak growth rate, peak growth time, doubling time (at the peak growth), lag time, max y, and max y time. View the results by calling each list element like:   

```{r view reults}
out.singlesample$estimates
```

For more instructions, please refer to the vignette.
