---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ipolygrowth

<!-- badges: start -->
<!-- badges: end -->

The goal of ipolygrowth is to calculate bacterial growth curve parameters using fourth degree polynomial functions. Functions are available for a single biological sample or multiple samples.


## Installation

The package can be installed from CRAN with:

```{r, eval = FALSE}
install.packages("ipolygrowth")
```

or the development version from [GitHub](https://github.com/) with:

``` {r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("https://github.com/kivanvan/ipolygrowth", upgrade = F, quiet = T)
```


## Example

Load the packages first.

```{r, message = FALSE}
library(ipolygrowth)
library(dplyr)
```

The example data comes from the [growthrates](https://CRAN.R-project.org/package=growthrates) package.  

```{r data}
# example data comes from the growthrates package (available on CRAN)
if (!"growthrates" %in% installed.packages()) {install.packages("growthrates")}
data <- growthrates::bactgrowth
```

Alternatively, download the `bactgrowth.txt` from [here](https://github.com/tpetzoldt/growthrates/tree/master/data) to the directory of your script. The data can then be read using the following code.

```{r alt data, eval = FALSE}
data <- read.table("bactgrowth.txt", header = TRUE) %>%
  mutate(strain = factor(strain, levels = c("D", "R", "T")))
```

This is a basic example which shows you how to use the single sample function:  

```{r example}
# subset data to a single biological sample
df.singlesample <- data %>% dplyr::filter(strain == "D", conc == 0)

# calculate growth curve parameters using ipolygrowth function
out.singlesample <- ipg_singlesample(data = df.singlesample, time.name = "time", y.name = "value")
```


The output is a list, including a table of growth parameter estimates, the polynomial model, a table of beta coefficients, and a table of fitted values. Growth parameters include peak growth rate, peak growth time, doubling time (at the peak growth), lag time, max y, and max y time. View the results by calling each list element like:   

```{r view reults}
out.singlesample$estimates
```

For more instructions and the expected output of vignette, please refer to the [vignette on CRAN](https://cran.r-project.org/web/packages/ipolygrowth/vignettes/ipolygrowth_vignette.html).
